---
editor_options:
  markdown:
    wrap: 72
output: pdf_document
---

# Final report

-   Group Topic: Politics and Policy

-   Team members: Ming DeMers, Danielle Donovan, Julian Kanu, Daisy
    Meng, Kate Sloan, Jemima Yoon

## Causal Question

##### Describe your causal question in a way that someone who has not taken this class would understand. Why are you interested in this question? How could answering this question allow for better decision making? Include any necessary background or context. Cite outside sources you use.

How might the distance that a constituent lives from their legislator
affect the likelihood of the constituent getting a response to a request
sent to their legislator.

This question is interesting because legislators oftentimes have to make
decisions in situations that are difficult to observe, e.g. behind
closed doors and in high security locations. In the context of this
question, it is difficult to observe a given legislator in deciding what
kind of requests from their constituents they will and will not respond
to. The location of the constituent’s residence can be a factor that
influences the legislator’s decision to respond to their constituents.
Moreover, we as constituents do not always pay attention to the types of
issues that legislators are presented with (and have to make a decision
on). This question has significant real-life implications; for example,
if legislators are found to be more likely to respond to the
constituents who live within the same district, this may encourage
higher voter turnout among the local constituents who feel like their
voices are being heard.

**Describe your causal question in the language of causal inference
we’ve learned in this course: What is the treatment? What is the
outcome? What are the potential outcomes? Write these out in words and
in the math notation we have used in class.**

$A_i = \begin{cases} 0 & \text{If location far} \\ 1 & \text{ If location is close} \end{cases}$

$Y_{constituent}^{a=1}$

$Y_{constituent}^{a=0}$

Treatment: Proximity of constituent

Outcome: Usefulness of the legislator's response to the constituent’s
letter.

##### Draw a DAG representing your causal question that includes at least three relevant variables besides treatment and outcome that are included in your dataset. You may include more than three variables. You may include variables that are not in your dataset, but at least 3 of your variables (excluding treatment and outcome) must be included in your dataset. If you use letters to denote variables, make sure they are clearly defined

**Answer**

##### Explain your DAG: tell us in words what is meant by each edge in your DAG.

A-\>Y This is a direct causal path for location and the usefulness of
the response. A\<-M-\>Y This shows how the median household income which
is M impacts the location in which you are living which also influences
the response that you get. R-\>P-\>Y: The amount of rural housing( R)
will influence the legislative party as rural areas tend to be redder,
and it will also affect the response as there can be more or fewer
incentives based on your party. A\<-P-\>Y: This shows how the
legislative party(P) of an area impacts the location as we are assuming
that red areas tend to be more rural, which also impacts the response
since some politicians may prefer to represent certain areas or
demographics.

##### Discuss your DAG. How realistic is it? Are there variables or edges you excluded from your DAG that someone else might argue should be included? Playing devil’s advocate, how would you critique the reliability of your DAG?

This DAG is realistic because the variables included are fairly broad,
and their effects on multiple other variables are demonstrated. Although
our study focuses on location and response, median household income and
the legislative party can impact both of these variables, as explained
above. We excluded race, however, which we initially considered a
confounder. However, our study chooses to focus on other factors that
may also be affected by race.. Race may affect location and response
outside of median household income because of other systemic bias or
deliberate manipulation we couldn’t formalize in our dag:systemic
segregation, redlining, and gerrymandering, for example. Another
variable that one may argue should be included is age; some politicians
make it a goal to listen to their younger constituents, while some
politicians' voter base may be older, which will impact the helpfulness
of the response.

## Method and Identification

##### What method are you using to estimate a causal effect? What causal effect are you estimating (ATE vs LATE vs ATT)? What assumptions are required to identify the causal effect via your chosen method?

We use nearest neighbor model as it favors a fast, but accurate result
over medium and large sample sizes. This approach matches treated units
(constituents close to the legislator) to control units (those farther
away) based on their similarity in observed covariates. The causal
effect is estimated by average treatment effect on treated (ATT). This
measures the group as a whole, and finding if they are better or worse
off given a treatment, on average.

Some of the assumptions that are made to  identify the causal effect via
ATT are the idea of conditional exchangeability. And we ultimately chose
ATT because we want to find the average treatment effect of the treated,
due to the design of the experiment we would only want to see the
outcome to those who actually specified their location(treatment) and
not those who do not match the the far or near criteria. 

##### Explain what conditional exchangeability means in the context of your causal question. Is it important? Why or why not? How do sufficient adjustment sets relate to conditional exchangeability?

M(Median income) and P(political Party)  are confounding variables and
need to be blocked. By blocking the backdoor paths A\<-M-\>Y and 
A\<-P-\>Y, we satisfy the conditional exchangeability assumption. This
is important because the location of constituents is not randomized; we
are working with observational data. As we have explained, location is
affected by political party and income, unless we condition on these
variables, we will not satisfy any type of exchangeability. If we do not
satisfy the exchangeability assumption, we cannot tell if the response
rate is caused by income or party rather than location. When we satisfy
conditional exchangeability, we can be more confident on the causal
effect of location.

##### Assuming your DAG is true, list out all non-causal paths between treatment and outcome and list one sufficient adjustment set to identify the causal effect of the treatment on the outcome. If a sufficient adjustment set does not exist, add additional variables to your DAG so that one does exist.

All non-causal paths in the DAG above include: 

A\<-M-\>Y, median household income impacts the location in which you are
living which also influences the response that you get. 

R-\>P-\>Y: Amount of rural housing will influence the legislative party
will also affect the response as there can be more or fewer incentives
based on your party. 

A\<-P-\>Y: This shows how the legislative party of an area impacts the
location as we are assuming that red areas tend to be more rural, which
also impacts the response.

One sufficient adjustment set is {P, M}. By adjusting for M and P, you
will see the isolated causal effect of A to Y since you would be
blocking the back door paths.

##### Discuss the plausibility of conditional exchangeability in your setting. If your sufficient adjustment set contains variables that are not in your dataset, discuss the implications.

The sufficient adjustment set does not have any variables that are not
found in the data set, as the dataset and its observations are
extensive. Because the data accounts for multiple confounders,
conditional exchangeability is plausible. Additionally, suppose we
adjust the sufficient adjustment set identified previously {P, M} and
assume that no other factors affect the effect of A to Y. In that case,
it can be assumed that the direct causal path from A to Y would be
legitimate. By controlling for political party and median household
income, we can assume that the most significant confounding variables
have been accounted for. Many smaller factors can be mostly accounted
for within these variables. 

As mentioned previously, it is possible that there are variables not in
the data set that could be confounders on their own, such as race
(included in data set) or age of the constituent (not included in the
data set). If these confounders are relevant, it could impact the
plausibility of conditional exchangeability. However, we have reasoned
that these variables are not relevant to our research since there is no
effort to identify or separate the age groups of the constituents, and
race would not necessarily affect distance.

##### Discuss any other identification assumptions for your method here, such as positivity and consistency. What do they mean in the context of your causal question and are they plausible?

Positivity may be violated in cases if only rural areas or low-income
areas align one party (like Republicans) and live very far away from the
location. However, positivity would not be violated if there are
variations in party alignment P or variations in M (median income) and
the location is close. We assume that the positivity would not be
violated as income and political alignment is not a determining factor
of whether or not one lives further away from the politician. Thus the
treatments exist. 

While we can assume that a certain type of people may choose to live
more rural or urban, there is not a clear cut-off. There is some
correlation between income and political party that separates people
into different neighborhoods based on both of these factors. However,
there is no reason to believe that neighborhoods are dispersed in such a
way that one group (high or low income/Republican or Democrat) would be
mostly either in the treated or untreated group. We are also assuming
that there is consistency in our method since the distance is already
defined in the data set. The distance is precise and divides everyone
into two exclusive categories. Y=Y\^a and A=a. It is also consistent
because there is clarity about interference. This is true because
people’s messages and responses should be independent from one another. 

However, one thing that might cause interference is that in some cases,
someone may be more likely to receive a response if their neighbor does.
If an issue is affectinga specific area and many people complain about
it, a resident in that group may be more likely to receive a response
than someone who makes a unique complaint.

## Discussion: Analysis and Results

##### Give some context for your dataset. Who is included in your dataset? How was the data collected? When was the data collected? Make sure to cite the dataset.

The dataset consists of demographic information regarding the
legislators, the districts they represent (including voter turnout and
urban/rural housing), whether or not the legislators responded to the
constituents’ requests, and whether or not the legislators were helpful
in their response. The original data on the legislators and their
districts involved in the experiment was collected through public state
legislative websites for all 50 states within the USA. This information
includes the legislators’ race, party affiliation, and gender, as well
as the racial diversity of the districts and the extent of urbanization
within each district. The researcher that collected this data did not
note when he collected this data, but given that his findings were
published in 2013, the data was most likely collected about 2-3 years
prior (2010-2012) (cleaned dataset:
<https://github.com/Ming-DeMers/3900-final/blob/main/data/poli1.Rdata>)
( original Data set:
<https://dataverse.harvard.edu/file.xhtml?persistentId=doi:10.7910/DVN/AMMI3S/QCV5DG&version=2.1&toolType=PREVIEW> 
)

##### Discuss any choices you made regarding data cleaning and processing: Did your data have missing values or outliers? How did you handle them? Were there any variables you dichotomized (i.e. made binary), or variables that you changed the format (e.g. yes/no to 1/0)?

The original dataset was already very well-organized, so there was not
much to clean. While there was one variable with missing values
(percentage of those in a legislator’s district who voted Democratic),
we were not interested in the variable and so the missing values did not
pose an issue. 

We considered making the distance binary and setting up a threshold to
what constitutes far or near. However, this dataset was constructed
where the treatment is already considered “faraway city” or “nearby
city,” and a column in the dataset indicating whether or not the
legislator received the treatment.Thus, we didn’t think the actual
quantified distances we have in the dataset are too important.
Ultimately we did not dichotomize or change the format of any variables.

##### Discuss the impact of any choices you made regarding your dataset, such as choices you made in data cleaning or processing.

Initially, we removed the urban/rural housing variables because we did
not think it would be included in the DAG. However, redecided when
designing the DAG. We reincluded the urban/housing variables. This is
because, depending on the area, living more urban or rural may impact
one's political affiliation which may also in turn, impact the
politician's response rate.

##### Explain how you estimated a causal effect.

-   If you used matching, explain and discuss your choices. What formula
    did you use and why? What matching strategy did you use and why? Are
    there any advantages or drawbacks to the strategy you chose? How
    many units did your matching drop? How was the covariate balance in
    your matched sample? Discuss the implications of any choices you
    made and the quality of your matching.
-   If you didn’t use matching, explain any choices you made related to
    the method you used and discuss their implications. Think about
    advantages or drawbacks to any choices you made, possible
    bias-variance trade-offs, and assessing how well your method did.

We estimated the causal effect of proximity on legislator responsiveness
using propensity score matching. The treatment variable, treat_out,
indicating proximity, was modeled as a function of covariates to ensure
matching based on all variables likely to influence both treatment
assignment (proximity) and the outcome (responsiveness), as shown by our
DAG, thus reducing confounding.

We used nearest-neighbor matching with replacement. This method pairs
treated units with the closest control units ensuring immediate matches
with minimal differences.Allowing replacement improved match quality,
especially when control units were limited, but also meant that some
control units were reused multiple times, potentially reducing
representativeness.

Matching dropped 1,393 control units, reflecting a potential positivity
violation. After matching, covariate balance improved substantially
between treated and control groups, suggesting the matching process was
effective in creating comparable groups.

The advantages include its ability to reduce bias from observed
confounders and to make causal estimates more credible. However, its
reliance on the previously mentioned assumptions on confounders and
exchangeability means that any unmeasured confounders could still bias
the estimate. Additionally, dropping unmatched units could limit the
generalizability of the findings to the matched sample. Overall, the
method provided a credible estimate of the ATT, but the quality of the
matching and potential for unobserved confounding highlight areas where
the analysis could be improved.

##### Report your causal effect estimate and interpret it in the context of your causal question.

Our estimated causal effect, the ATT, is -0.12. This means proximity, on
average, decreased the likelihood of a legislator responding to a
request by approximately 12 points. That is, for our constituents that
are closer to their legislator, they are less likely to receive a
response to a request that than those who live further away, by around
0.12%

This result suggests that proximity might negatively influence
responsiveness, possibly due to factors like higher demand from nearby
constituents, strategic political behavior, or other unobserved
confounders associated with being "close." One possible mathematical
explanation would be that, given a steady density of inhabitants at any
given distance from the legislator, there will be more total inhabitants
in the circumference of a larger radius than a smaller one.

Overall, the causal effect estimate suggests that simply being
geographically closer to a legislator does not guarantee greater
responsiveness and, in this case, appears to reduce it.

##### Discuss the limitations of your analysis: what are the limitations of your dataset? Is there other data you would have wanted to have to bolster your analysis? Playing devil’s advocate, how would you critique the reliability of you causal estimate?

Our analysis is ultimately limited by the nature of the dataset and the
assumptions required for the methodology. Foremost, the dataset is
synthetic, meaning it is constructed to simulate real-world scenarios,
yet may not fully capture the complexities of legislative behavior and
constituent-legislator dynamics.

The binary treatment variable (treat_out), which indicates proximity,
oversimplifies distance into a treated or control category. A continuous
measure of proximity could’ve better capture the nuances of how distance
affects responsiveness. However, we would have binarized any linear
measurement as otherwise would’ve been outside the scope of this
project. 

To bolster the analysis, we would have liked to include data on the
nature and content of constituent requests (e.g., their urgency or
importance), as this likely influences whether a legislator responds.
What are the actual contents of the requests? Legislators likely have an
agenda they focus on for each term, and thus targe certain requests more
than others, despite the distance of the residence of requester.
Legislative factors such as staff capacity, workload, and district size
could also provide a clearer understanding of constraints on
responsiveness. 

Critiquing the causal estimate, the analysis relies on several key
assumptions, positivity and consistency, and ultimately conditional
exchangeability. If unmeasured factors, like the complexity of
constituent issues or district-specific political dynamics, influence
both treatment and outcome, the estimate may be biased. While propensity
score matching improved balance, residual imbalances in covariates and
the exclusion of 1,393 unmatched control units suggest potential
violations of the positivity assumption. This could restrict the
generalizability of the findings to the matched sample rather than the
broader population.

Moreover, the lower r-squared value in the regression model could
indicate that only a small proportion of the variance in responsiveness
is explained by the covariates included, suggesting other unobserved
factors may play a significant role. While the analysis provides
insights into how proximity might influence responsiveness, we can, at
best, say our conclusions apply to this synthetic dataset, and may apply
to a real-world example. But more data collection and research would be
necessary.

\newpage

# Code:

```{r}
# This is where your code goes for the data cleaning/processing and analysis. 
# Make your code clean and easy-to-follow. Add short comments to explain what you are doing. 
# If a classmate who wasn’t as familiar with R were to read through this section, 
# would they be able to follow along?

library(tidyverse)
library(MatchIt)
```

## Data Cleaning

```{r eval=F}
#|label: dataset-clean

# Cleaning data 
View(x) # x was automatically named while saving the original dataset 
poli <- x # renamed the dataset
poli <- poli[ -c(23,24:25) ] # removed a few columns I deemed unnecessary
poli <- poli[-c(28)]
poli[-c(31)]
poli <- cbind(poli, x['ruralhousing']) # readded this column and the following from the original dataset after discussing what variables my team members and I wanted to include in our DAG

poli <- cbind(poli, x['ruralhousing'])
poli <- cbind(poli, x['urbanhouses'])
poli <- cbind(poli, x['totalhouses'])

save(poli, file = "data/poli1.Rdata") # saved the file to my computer 

```

## Matching

We attempt to match.

```{r}
#|label: matching

load("data/poli1.Rdata")


# create matches based off the treatment (treat_out) and outcome
# (code_some_response), along with possible covariates leg_party and medianhhincome
# matching is done using the nearest neighbor method (method = "nearest"`)
# with a logistic regression model for estimating propensity scores (distance = "logit")
# the causal effect measurement is by ATT (estimand) = "ATT").
# `ratio = 1` specifies 1-to-1 matching, and `replace = T` allows controls to be reused
match <- matchit(treat_out ~ code_some_response_given + leg_party + medianhhincom,
                 data = poli,
                 method = "nearest",
                 distance = "logit",
                 estimand = "ATT",
                 ratio = 1,
                 replace = T)

# extract the matched dataset resulting from the propensity score matching
match.data(match) -> matched_df

# fit a linear regression model on the matched dataset
# this model estimates the treatment effect (causal effect) of treat_out
lm_mod <- lm(code_some_response_given ~ treat_out + leg_party + medianhhincom,
             data = matched_df)

# display a summary of the regression model, including coefficient estimates, 
# standard errors, and statistical significance
summary(lm_mod)

# extract the estimated treatment effect (ATT) from the regression model,
# and print the actual causal estimate
att_est <- coef(lm_mod)["treat_out"]
print(att_est)

# show how many units matched and unmatched
summary(match)$nn # number of matched

# print the causal estimate (ATT) rounded to 3 places

print(round(att_est, 3)) 
```
